### 2024-06-02 2024-5-13/sections/algorithms.tex
commit the paper extended by setta2023
change nonemptiness to non-emptiness

\begin{proposition}[\cite{atva2020}]\label{prop-cefadec}
$\cefadec$ is PSPACE-complete. 
\end{proposition}
%\begin{proof}
%  The lower bound follows from the fact that the nonemptiness problem of the intersection of NFAs is PSPACE-complete \cite{pspace_int}. For the upper bound, \cite{atva2020} showed that $\cefadec$ is in PSPACE by reducing it to monotonic counter machines.
%\end{proof}
Note that the decision procedure in \cite{atva2020} was only used to prove the upper bound in Proposition~\ref{prop-cefadec} and not implemented as a matter of fact. Instead, the symbolic model checker nuXmv~\cite{nuxmv} was used to solve $\cefadec$. We do not rely on nuXmv in this work and shall propose a new algorithm for solving $\cefadec$ in Section~\ref{subsec:cefadec}.

@@

At first, we show how to construct a CEFA from a regex where counting operators may occur. 
Let us start with register-representable regexes defined in the sequel.

@@

For each $i \in [n]$, let $x_i \in e_{i, 1}, \cdots, x_i \in e_{i, l_i}$ be an enumeration of the regular membership constraints for $x_i$ in $\varphi_1$.  Then we can construct CEFA $\aut_{i, j}$ from $e_{i, j}$ for each $i \in [n]$ and $j \in [l_i]$. Moreover, we construct another CEFA $\aut_{i, 0}$ for each $i \in [n]$ to model the length of $x_i$. Specifically, $\aut_{i,0}$ is constructed as $((r_{i,0}), \{q_{i,0}\}, \Sigma, \delta_{i,0}, \{q_{i,0}\}, \{q_{i,0}\}, \ltrue)$ where $r_{i,0}$ is a fresh register and $\delta_{i,0} = \{(q_{i,0}, a, q_{i,0}, (1)) \mid a \in \Sigma\}$. 
Let $\Lambda_{x_i} = \{\aut_{i,0}, \aut_{i, 1}, \cdots, \aut_{i, l_i}\}$ for each $i \in [n]$, and $\varphi'_2 \equiv \varphi_2[r_{1,0}/|x_1|, \cdots, r_{n,0}/|x_n|]$. Then the satisfiability of $\varphi$ is reduced to the nonemptiness of CEFAs in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ w.r.t. $\varphi'_2$.

@@

After Step 1, the nonemptiness of CEFAs in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ w.r.t. $\varphi$ is reduced to the nonemptiness of CEFAs in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$.

@@

Therefore, to solve the nonemptiness of CEFA in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$, it is sufficient to compute the existential LIA formulas $\psi_{\cU_{\cB_1}}(\fZ_{\costset_{\cB_1}}), \cdots, \psi_{\cU_{\cB_n}}(\fZ_{\costset_{\cB_n}})$ to represent the Parikh images of $\cU_{\cB_1}$, $\cdots$, $\cU_{\cB_n}$ respectively, where $\fZ_{\costset_{\cB_i}} = \{\anivarz_{i,\myvec{v}} \mid \myvec{v} \in \costset_{\cB_i}\}$ for $i \in [n]$, and solve the satisfiability of the following existential LIA formula
\[
\begin{array}{l}
\bigwedge \limits_{i \in [n]} \left(\psi_{\cU_{\cB_i}}(\fZ_{\costset_{\cB_i}}) \wedge \alpha_i\left[\sum\limits_{\myvec{v} \in \costset_{\cB_i}} \anivarz_{i, \myvec{v}} \myvec{v} \big{/}R_{\cB_i}  \right] \right)\  \wedge \\
\ \ \varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cB_1}} \anivarz_{1, \myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cB_n}} \anivarz_{n, \myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
\end{array}
\]
Intuitively, the integer variables $\anivarz_{i, \myvec{v}}$ represent the number of occurrences of $\myvec{v}$ in the strings accepted by $\cU_{\cB_i}$.

>>>>

\begin{proposition}[\cite{atva2020}]\label{prop-cefadec}
$\cefadec$ is PSPACE-complete. 
\end{proposition}
%\begin{proof}
%  The lower bound follows from the fact that the non-emptiness problem of the intersection of NFAs is PSPACE-complete \cite{pspace_int}. For the upper bound, \cite{atva2020} showed that $\cefadec$ is in PSPACE by reducing it to monotonic counter machines.
%\end{proof}
Note that the decision procedure in \cite{atva2020} was only used to prove the upper bound in Proposition~\ref{prop-cefadec} and not implemented as a matter of fact. Instead, the symbolic model checker nuXmv~\cite{nuxmv} was used to solve $\cefadec$. We do not rely on nuXmv in this work and shall propose a new algorithm for solving $\cefadec$ in Section~\ref{subsec:cefadec}.

@@

At first, we show how to construct a CEFA from a regex where counting operators may occur. Our approach is based on Thompson's construction~\cite{Thompson1968} for regexes without counting operators. Let us start with register-representable regexes defined in the sequel.

@@

For each $i \in [n]$, let $x_i \in e_{i, 1}, \cdots, x_i \in e_{i, l_i}$ be an enumeration of the regular membership constraints for $x_i$ in $\varphi_1$.  Then we can construct CEFA $\aut_{i, j}$ from $e_{i, j}$ for each $i \in [n]$ and $j \in [l_i]$. Moreover, we construct another CEFA $\aut_{i, 0}$ for each $i \in [n]$ to model the length of $x_i$. Specifically, $\aut_{i,0}$ is constructed as $((r_{i,0}), \{q_{i,0}\}, \Sigma, \delta_{i,0}, \{q_{i,0}\}, \{q_{i,0}\}, \ltrue)$ where $r_{i,0}$ is a fresh register and $\delta_{i,0} = \{(q_{i,0}, a, q_{i,0}, (1)) \mid a \in \Sigma\}$. 
Let $\Lambda_{x_i} = \{\aut_{i,0}, \aut_{i, 1}, \cdots, \aut_{i, l_i}\}$ for each $i \in [n]$, and $\varphi'_2 \equiv \varphi_2[r_{1,0}/|x_1|, \cdots, r_{n,0}/|x_n|]$. Then the satisfiability of $\varphi$ is reduced to the non-emptiness of CEFAs in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ w.r.t. $\varphi'_2$.

@@

After Step 1, the non-emptiness of CEFAs in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ w.r.t. $\varphi$ is reduced to the non-emptiness of CEFAs in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$.

@@

Therefore, to solve the non-emptiness of CEFA in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$, it is sufficient to compute the existential LIA formulas $\psi_{\cU_{\cB_1}}(\fZ_{\costset_{\cB_1}}), \cdots, \psi_{\cU_{\cB_n}}(\fZ_{\costset_{\cB_n}})$ to represent the Parikh images of $\cU_{\cB_1}$, $\cdots$, $\cU_{\cB_n}$ respectively, where $\fZ_{\costset_{\cB_i}} = \{\anivarz_{i,\myvec{v}} \mid \myvec{v} \in \costset_{\cB_i}\}$ for $i \in [n]$, and solve the satisfiability of the following existential LIA formula
\[
\begin{array}{l}
\bigwedge \limits_{i \in [n]} \left(\psi_{\cU_{\cB_i}}(\fZ_{\costset_{\cB_i}}) \wedge \alpha_i\left[\sum\limits_{\myvec{v} \in \costset_{\cB_i}} \anivarz_{i, \myvec{v}} \myvec{v} \big{/}R_{\cB_i}  \right] \right)\  \wedge \\
\ \ \varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cB_1}} \anivarz_{1, \myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cB_n}} \anivarz_{n, \myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
\end{array}
\]
Intuitively, the integer variables $\anivarz_{i, \myvec{v}}$ represent the number of occurrences of $\myvec{v}$ in the strings accepted by $\cU_{\cB_i}$.

### 2024-06-02 2024-5-13/sections/overview.tex
change nonemptiness to non-emptiness

Finally, the satisfiability of the original string constraint is reduced to the nonemptiness of the CEFA $\aut\equiv \aut_1 \cap \aut_2 \cap \aut_3$ with respect to the LIA formula $\varphi \equiv r_4 > 120$, that is, whether there exist $w \in \Sigma^*$ and $(v_1, v_2, v_3, v_4) \in \Int^4$ such that $w$ is accepted by $\aut$, so that the resulting registers values $(v_1, v_2, v_3, v_4)$ satisfy both $1 \le v_1 \le 60\wedge 1 \le v_2 \le 60 \wedge 0 \le v_3 \le 60$ and $\varphi$. 
It is not hard to observe that the nonemptiness of $\aut$ with respect to $\varphi$ is independent of the characters of $\aut$.  Therefore, the characters in $\aut$ can be ignored, resulting into an NFA $\cB$ over the alphabet $\costset$, where $\costset$ is the set of vectors from $\Int^4$ occurring in the transitions of $\aut$ (see Figure~\ref{fig:overview:product:reduced}(a)). Then the original problem is reduced to the problem of deciding whether there exists a string $w' \in \costset^*$ that is accepted by $\cB$ and its Parikh image (i.e., numbers of occurrences of characters), say $\eta_{w'}: \costset \rightarrow \Nat$, satisfies $1 \le v'_1 \le 60\wedge 1 \le v'_2 \le 60 \wedge 0 \le v'_3 \le 60 \wedge v'_4 > 120$, where $(v'_1, v'_2, v'_3, v'_4) =  \sum \limits_{\myvec{v} \in \costset} \eta_{w'}(\myvec{v}) \myvec{v}$ for each $\myvec{v}\in \costset$. Intuitively, $(v'_1, v'_2, v'_3, v'_4)$ is a weighted sum of vectors $\myvec{v} \in \costset$, where the weight is the number of occurrences of $\myvec{v}$ in $w'$. (See Section~\ref{subsec:cefadec} for more detailed arguments.)

>>>>

Finally, the satisfiability of the original string constraint is reduced to the non-emptiness of the CEFA $\aut\equiv \aut_1 \cap \aut_2 \cap \aut_3$ with respect to the LIA formula $\varphi \equiv r_4 > 120$, that is, whether there exist $w \in \Sigma^*$ and $(v_1, v_2, v_3, v_4) \in \Int^4$ such that $w$ is accepted by $\aut$, so that the resulting registers values $(v_1, v_2, v_3, v_4)$ satisfy both $1 \le v_1 \le 60\wedge 1 \le v_2 \le 60 \wedge 0 \le v_3 \le 60$ and $\varphi$. 
It is not hard to observe that the non-emptiness of $\aut$ with respect to $\varphi$ is independent of the characters of $\aut$.  Therefore, the characters in $\aut$ can be ignored, resulting into an NFA $\cB$ over the alphabet $\costset$, where $\costset$ is the set of vectors from $\Int^4$ occurring in the transitions of $\aut$ (see Figure~\ref{fig:overview:product:reduced}(a)). Then the original problem is reduced to the problem of deciding whether there exists a string $w' \in \costset^*$ that is accepted by $\cB$ and its Parikh image (i.e., numbers of occurrences of characters), say $\eta_{w'}: \costset \rightarrow \Nat$, satisfies $1 \le v'_1 \le 60\wedge 1 \le v'_2 \le 60 \wedge 0 \le v'_3 \le 60 \wedge v'_4 > 120$, where $(v'_1, v'_2, v'_3, v'_4) =  \sum \limits_{\myvec{v} \in \costset} \eta_{w'}(\myvec{v}) \myvec{v}$ for each $\myvec{v}\in \costset$. Intuitively, $(v'_1, v'_2, v'_3, v'_4)$ is a weighted sum of vectors $\myvec{v} \in \costset$, where the weight is the number of occurrences of $\myvec{v}$ in $w'$. (See Section~\ref{subsec:cefadec} for more detailed arguments.)
